---
title: "Hands-on Exercise 3: Processing and Visualizing Flow Data"
author: "Kristine Joy Paas"
date: "1 December 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

We will visualize flow data

::: {.callout-caution appearance="simple"}
I am running through most of the code here as the wrangling steps are very similar to the wrangling done in Take Home Ex1. No need to analyze each and every step as I already learned those.
:::

## Getting Started

We will load the library used for this exercise.

```{r}
pacman::p_load(tmap, sf, DT, stplanr,
               performance,
               ggpubr, tidyverse)
```

# Preparing flow data

## Importing of data

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202310.csv")
```

```{r}
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE) 
```

```{r}
head(odbus)
```

## **Extracting the study data**

```{r}
odbus6_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 6 &
           TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE,
           DESTINATION_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

```{r}
datatable(odbus6_9)
```

```{r}
write_rds(odbus6_9, "data/rds/odbus6_9.rds")
odbus6_9 <- read_rds("data/rds/odbus6_9.rds")
```

# **Working with Geospatial Data**

## **Importing geospatial data**

```{r}
busstop <- st_read(dsn = "data/geospatial",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                   layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

# **Geospatial data wrangling**

## **Combining Busstop and mpsz**

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>%
  select(BUS_STOP_N, SUBZONE_C) %>%
  st_drop_geometry()
```

```{r}
datatable(busstop_mpsz)
```

```{r}
write_rds(busstop_mpsz, "data/rds/busstop_mpsz.rds")  
```

```{r}
od_data <- left_join(odbus6_9 , busstop_mpsz,
            by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE,
         ORIGIN_SZ = SUBZONE_C,
         DESTIN_BS = DESTINATION_PT_CODE)
```

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
```

```{r}
od_data <- unique(od_data)
```

```{r}
od_data <- left_join(od_data , busstop_mpsz,
            by = c("DESTIN_BS" = "BUS_STOP_N")) 
```

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
od_data <- unique(od_data)
```

```{r}
od_data <- od_data %>%
  rename(DESTIN_SZ = SUBZONE_C) %>%
  drop_na() %>%
  group_by(ORIGIN_SZ, DESTIN_SZ) %>%
  summarise(MORNING_PEAK = sum(TRIPS))
```

```{r}
write_rds(od_data, "data/rds/od_data.rds")
od_data <- read_rds("data/rds/od_data.rds")
```

# **Visualizing Spatial Interaction**

## **Removing intra-zonal flows**

```{r}
od_data1 <- od_data[od_data$ORIGIN_SZ!=od_data$DESTIN_SZ,]
```

## **Creating desire lines**

In this code chunk below, `od2line()` of **stplanr** package is used to create the desire lines.

```{r}
flowLine <- od2line(flow = od_data1, 
                    zones = mpsz,
                    zone_code = "SUBZONE_C")
```

## **Visualizing the desire lines**

To visualise the resulting desire lines, the code chunk below is used.

```{r}
tm_shape(mpsz) +
  tm_polygons() +
flowLine %>%  
tm_shape() +
  tm_lines(lwd = "MORNING_PEAK",
           style = "quantile",
           scale = c(0.1, 1, 3, 5, 7, 10),
           n = 6,
           alpha = 0.5,
           col = "red")
```

```{r}
tm_shape(mpsz) +
  tm_polygons() +
flowLine %>%  
  filter(MORNING_PEAK >= 5000) %>%
tm_shape() +
  tm_lines(lwd = "MORNING_PEAK",
           style = "quantile",
           scale = c(0.1, 1, 3, 5, 7, 10),
           n = 6,
           alpha = 0.5,
           col = "red")
```

# Reflections

-   While doing the exercise, I was interested in doing analysis of the flow from origin to destination to figure out the movement patterns of people. However, we do not know where people are going so even when I had hypothesis like people come from the same place in mornings (their home), whether weekend or weekday while it's not true in the evenings as they come from school or office, I couldn't verify it with LISA and EHSA they analyze only 1 data parameter (like TRIPS) in Take home ex 1

-   I am too tired so I didn't have much time and energy to work on this (because of Take Home EX 1. Good thing the exercise is shorter and I already understand the data wrangling part so I can focus more on the flow analysis.
