---
title: "Take Home Exercise 1C: Spatial Analysis with LISA"
author: "Kristine Joy Paas"
date: "29 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

The aim of this study is to uncover spatial and spatio-temporal mobility patterns of public bus passengers in Singapore.

The main modes of analysis to be used here are Local Indicators of Spatial Association (GLISA) and Emerging Hot Spot Analysis (EHSA).

In doing these study, we will be looking at bus trips started during the hours below.

| Peak hour period             | Bus tap on time |
|------------------------------|-----------------|
| Weekday morning peak         | 6am to 9am      |
| Weekday evening peak         | 5pm to 8pm      |
| Weekend/holiday morning peak | 11am to 2pm     |
| Weekend/holiday evening peak | 4pm to 7pm      |

More details about the study can be found [here](https://isss624-ay2023-24nov.netlify.app/take-home_ex01).

In this part of the study, we will do spatial analysis with LISA using bus commuter traffic data generated from [Data Wrangling](/Take-home_Ex1/Take-home_Ex1A.qmd). We will also attempt the answer the [Open Questions](/Take-home_Ex1/Take-home_Ex1B.html#conclusion) from [Geovisualization and Analysis](/Take-home_Ex1/Take-home_Ex1B.html):

-   Which areas are busier on weekends? on weekdays?

-   What are the commuting patterns of people during weekdays? weekends?

# Setup

## Setting Up the R Environment

We will load the following R packages needed for this study.

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html): for thematic mapping

-   [**sf**](https://cran.r-project.org/web/packages/sf/index.html): for geospatial data handling

-   [**tidyverse**](https://cran.r-project.org/web/packages/tidyverse/index.html): for non-spatial data handling

-   [**sfdep**](https://sfdep.josiahparry.com/): for spatial analysis

-   [**knitr**](https://cran.r-project.org/web/packages/knitr/):for prettifying presentation

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse, knitr)
```

## Environment settings

We will also set the default settings on for this document

-   `tmap_style` to **natural**: for displaying the maps with preferred style

-   set **seed** for reproducibility of results

```{r}
tmap_mode("plot")
tmap_style("natural")
set.seed(1234)
```

## Loading the data

::: callout-important
Before running this part, please run all the code chunks in [Data Wrangling](/Take-home_Ex1/Take-home_Ex1A.qmd) as it generates the data needed for this document.
:::

Use `read_rds()` to load the `rds` data needed for Local Indicator of Spatial Association (LISA) Analysis

```{r}
mpsz <- read_rds("data/rds/mpsz.rds")
honeycomb <- read_rds("data/rds/honeycomb202310.rds")
peak_trips_sf <- read_rds("data/rds/peak_trips_sf202310.rds")
```

-   `honeycomb` - contains the geometry of honeycomb grid containing Singapore bus stops
-   `peak_trips_sf` - contains number of bus trips originating from each hexagon for each peak period

# Methodology

Local Indicator of Spatial Association (LISA) detects spatial clusters present in our geospatial data and also helps identify spatial outliers.

We will use this analysis method to identify which areas have the highest bus commuters and which ones have the least.

To perform a LISA Analysis, we will do the following:

-   Deriving contiguity weights using Queen's method

-   Computing global Moran's I

-   Computing local Moran's I

-   Visualizing local Moran's I and p-value

-   Visualizing LISA Map

-   Interpreting the results

We will do this analysis for the all the peak hours.

::: callout-important
For all analysis in this study, we will set the significance level, or $\alpha$ value to **0.05**.
:::

# Computing adaptive distance weights matrix

Before getting Moran's I, we need to get the spatial weights first. We will use **adptive distance-based weights** for this, which uses the **k-nearest neighbors**.

::: {.callout-note collapse="true" appearance="simple"}
## Why not contiguity weights?

Visualizing the honeycomb grid reveals that there are **hexagon islands**, or **hexagons without contiguity neighbors**.

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
tm_shape(mpsz) +
  tm_fill("green", title = "Singapore Boundary", alpha = 0.5) +
  tm_shape(peak_trips_sf) +
  tm_fill(col = "white", title = "Hexagons", alpha = 1) +
  tm_borders(alpha = 0.2) +
  tm_layout(main.title = "Honeycomb grid corresponding to Singapore bus stops",
            main.title.position = "center",
            main.title.size = 1.0,
            legend.height = 0.35, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_compass(type="8star", size = 2, bg.color = "white", bg.alpha = 0.5) +
  tm_scale_bar(bg.color = "white", bg.alpha = 0.5) +
  tm_grid(alpha = 0.2)
```

Some of these are at the **East side of Singapore**, around **Changi Airport**.

We cannot associate them to another hexagon if we use contiguity neighbors so it will be difficult to do the analysis. With that, we will use distance-based weights to **ensure that all hexagons have neighbors**.
:::

As we are working with hexagons, which has **6 sides**, it is logical to get the **6 nearest neighbors** and use it for our analysis.

```{r}
knn6_nb <- peak_trips_sf %>% st_centroid() %>% st_knn(k=6)
head(knn6_nb, n = 3)
```

::: {.callout-note appearance="minimal"}
`st_centroid()` - used to get the centroids of each hexagon, which is required for `st_knn()`

`st_knn()` - used to get the k-nearest neighbors
:::

```{r}
wm_knn6 <- peak_trips_sf %>%
  mutate(
    nb = knn6_nb,
    wt = st_weights(nb, style = "W")
    )
```

# Calculating Global Moran's I

To check for signs of clustering, we will first calculate the global Moran's I value.

-   When $I > 0$, observations are similar (sign of clustering)

-   When $I < 0$, observations are dissimilar (low indication of clustering)

-   If $p < \alpha$ (**0.05**)**,** the result is significant and **did not happen by chance**

We will do this for all peak periods.

To perform the permutation test, we will use `global_moran_perm()` to perform Monte-Carlo simulations.

::: panel-tabset
## Weekday morning

```{r}
global_moran_perm(
  wm_knn6$WEEKDAY_AM_TRIPS,
  wm_knn6$nb,
  wm_knn6$wt,
  nsim = 99
)
```

-   $0.21487 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

## Weekday evening

```{r}
global_moran_perm(
  wm_knn6$WEEKDAY_PM_TRIPS,
  wm_knn6$nb,
  wm_knn6$wt,
  nsim = 99
)
```

-   $0.057327 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

## Weekend morning

```{r}
global_moran_perm(
  wm_knn6$WEEKEND_AM_TRIPS,
  wm_knn6$nb,
  wm_knn6$wt,
  nsim = 99
)
```

-   $0.15955 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

## Weekend evening

```{r}
global_moran_perm(
  wm_knn6$WEEKEND_PM_TRIPS,
  wm_knn6$nb,
  wm_knn6$wt,
  nsim = 99
)
```

-   $0.10265 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant
:::

::: {.callout-tip appearance="simple"}
## Insights

In all peak periods, $I > 0$ and $p < \alpha$ (**0.05**) so **there are signs of geospatial clustering** in the data.
:::

# Calculating Local Moran's I

As we have verified the existence of clusters, we can now proceed to looking for those clusters.

::: panel-tabset
## Weekday morning

```{r}
lmi_wkdy_am <- wm_knn6 %>% 
  mutate(local_moran = local_moran(
    WEEKDAY_AM_TRIPS, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

## Weekday evening

```{r}
lmi_wkdy_pm <- wm_knn6 %>% 
  mutate(local_moran = local_moran(
    WEEKDAY_PM_TRIPS, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

## Weekend morning

```{r}
lmi_wknd_am <- wm_knn6 %>% 
  mutate(local_moran = local_moran(
    WEEKDAY_AM_TRIPS, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

## Weekend evening

```{r}
lmi_wknd_pm <- wm_knn6 %>% 
  mutate(local_moran = local_moran(
    WEEKDAY_AM_TRIPS, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```
:::

# Plotting Local Moran's I and p-value

::: {.callout-note collapse="true" appearance="simple"}
## Function for plotting *ii* and *p_ii_sim* maps side by side

We will create a function to plot the local Moran's I and p-values side by side. This will allow for easy comparison of the 2 maps.

```{r}
plot_lmi_pv <- function(lmi) {
  tmap_arrange(
    tm_shape(mpsz) +
      tm_fill(col="green") +
      tm_borders(alpha = 0.5) +
      tm_shape(lmi) +
      tm_polygons("ii") + 
      tm_borders(alpha = 0.5),
    tm_shape(mpsz) +
      tm_fill(col="green") +
      tm_borders(alpha = 0.5) +
      tm_shape(lmi) +
      tm_polygons("p_ii_sim",
              breaks = c(0, 0.001, 0.01, 0.05, 1),
              labels = c("0.001", "0.01", "0.05", "Not sig"),
              palette = "YlOrRd"
              ) + 
      tm_borders(alpha = 0.5),
    ncol = 2
  )
}
```
:::

::: panel-tabset
## Weekday morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
plot_lmi_pv(lmi_wkdy_am)
```

## Weekday evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
plot_lmi_pv(lmi_wkdy_pm)
```

## Weekend morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
plot_lmi_pv(lmi_wkdy_pm)
```

## Weekend evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
plot_lmi_pv(lmi_wkdy_pm)
```
:::

::: {.callout-tip appearance="simple"}
## Insights

Big part of the maps show that most of the hexagons' p-values resulted into **insignificant values**. This means that those areas in red **do not have clusters**.

Visually, they all look similar. However, we can see the clusters and outliers more clearly once we plot the LISA maps.
:::

# Plotting LISA Maps

We will plot the Local Indicator of Spatial Association (LISA) maps to pinpoint where the clusters and outliers are.

To do that, we will only include those hexagons for which the results are significant, or `p_ii_sim` is less than our $\alpha$ of **0.05**.

::: panel-tabset
## Weekday morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"

tm_shape(mpsz) +
  tm_fill(col="white")+
  tm_borders(alpha = 0.5) +
  tm_shape(lmi_wkdy_am %>% filter(p_ii_sim < 0.05)) +
  tm_polygons("mean") + 
  tm_layout(
    main.title = "LISA for Weekday 6 - 9 AM",
    main.title.position = "center",
    main.title.size = 1
  )
```

::: {.callout-tip appearance="simple"}
### Insights

High-high areas can be observed at the **Causeway** bus stops. This can be attributed to **workers living in Johor Bahru commuting to Singapore for work** causing this bus stops to be busy in the mornings**.** Larger **high-high clusters** during this peak period. Most of them seem to correspond to **bus interchanges**.

Some high-low bus stops are can also be observed at the **West**. This could be key transfer points in these areas for people going to work or school.
:::

## Weekday evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"

tm_shape(mpsz) +
  tm_fill(col="white")+
  tm_borders(alpha = 0.5) +
  tm_shape(lmi_wkdy_pm %>% filter(p_ii_sim < 0.05)) +
  tm_polygons("mean") + 
  tm_layout(
    main.title = "LISA for Weekday 5 - 8 PM",
    main.title.position = "center",
    main.title.size = 1
  )
```

::: {.callout-tip appearance="simple"}
### Insights

Compared to weekday mornings, there are **smaller clusters in the Central** part of Singapore. This could indicate that traffic is more distributed in this area and people riding buses are **more spread out**.

One surprising observation is that the **Causeway bus stops** are not in a cluster during this peak period. It could indicate either lower traffic or more spread out traffic. One reason could be there are less people coming into the country during this time period.
:::

## Weekend morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"

tm_shape(mpsz) +
  tm_fill(col="white")+
  tm_borders(alpha = 0.5) +
  tm_shape(lmi_wknd_am %>% filter(p_ii_sim < 0.05)) +
  tm_polygons("mean") + 
  tm_layout(
    main.title = "LISA for Weekend/Holiday 11 AM - 2 PM",
    main.title.position = "center",
    main.title.size = 1
  )
```

::: {.callout-tip appearance="simple"}
### Insights

The clusters here coincide the most with **weekday morning** clusters. This could indicate that people are riding the bus from the same place, like their home.

Some notable differences are there is no cluster on the **Causeway bus stops** (less workers coming in to Singapore), and a **high-low outlier** in the **Southern** part. It is in close proximity to **NUS Hospital and Kent Ridge Park**. so perhaps people go for their morning run or visit the hospital.
:::

## Weekend evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
tm_shape(mpsz) +
  tm_fill(col="white")+
  tm_borders(alpha = 0.5) +
  tm_shape(lmi_wknd_pm %>% filter(p_ii_sim < 0.05)) +
  tm_polygons("mean") + 
  tm_layout(
    main.title = "LISA for Weekend/Holiday 4- 7 PM",
    main.title.position = "center",
    main.title.size = 1
  )
```

::: {.callout-tip appearance="simple"}
The **high-low** outlier in the South part of Singapore that is present in **weekend morning** is also present here. This is a potential recreation area or a destination for weekend errands/

Another key feature is the **triangle of high-low outliers** **in the West** (around Tuas and Jurong area). This pattern is also present in **weekday morning,** and **weekend morning** peak hours. These strongly supports the idea that these bus stops could be key transfer bus stops in this area.

In addition, **high-high clusters of 3 or more hexagons** are present in the same areas but are not present in the **weekday evening** peak hours. This can indicate similar behavior of people commuting from their homes.

Lastly, the **Causeway bus stops** are in a **high-high cluster** again, same as in **weekday morning**. This can indicate people coming back to Singapore after a short trip to Malaysia over the weekend.
:::
:::

::: {.callout-tip appearance="simple"}
## Insights

There are low-low clusters in the **West and Northwest** parts of the country. This is consistent with the previous observations that these areas have less traffic. In addition, other low-low clusters were also observed at the edges of Singapore like in the East and South. This could suggest that people in these areas **use alternative modes of transportation** like the MRT.

Another key observation is that the **low-high clusters** are in close proximity to the **high-high clusters**. On the other hand, **high-low** clusters are usually on their own. This could indicate that the bus stops in the low-high cluster are **under-utilized** because people prefer to use the bus stops in the nearby **high-high areas** there are **less bus routes** passing through these stops.
:::
