---
title: "Take Home Exercise 1C: Spatial Analysis with LISA"
author: "Kristine Joy Paas"
date: "29 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

The aim of this study is to uncover spatial and spatio-temporal mobility patterns of public bus passengers in Singapore.

The main modes of analysis to be used here are Local Indicators of Spatial Association (GLISA) and Emerging Hot Spot Analysis (EHSA).

In doing these study, we will be looking at bus trips started during the hours below.

| Peak hour period             | Bus tap on time |
|------------------------------|-----------------|
| Weekday morning peak         | 6am to 9am      |
| Weekday evening peak         | 5pm to 8pm      |
| Weekend/holiday morning peak | 11am to 2pm     |
| Weekend/holiday evening peak | 4pm to 7pm      |

More details about the study can be found [here](https://isss624-ay2023-24nov.netlify.app/take-home_ex01).

In this part of the study, we will do spatial analysis with LISA using bus commuter traffic data generated from [Data Wrangling](/Take-home_Ex1/Take-home_Ex1A.qmd). We will also attempt the answer the [Open Questions](/Take-home_Ex1/Take-home_Ex1B.html#conclusion) from [Geovisualization and Analysis](/Take-home_Ex1/Take-home_Ex1B.html):

-   Which areas are busier on weekends? on weekdays?

-   What are the commuting patterns of people during weekdays? weekends?

# Setup

## Setting Up the R Environment

We will load the following R packages needed for this study.

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html): for thematic mapping

-   [**sf**](https://cran.r-project.org/web/packages/sf/index.html): for geospatial data handling

-   [**tidyverse**](https://cran.r-project.org/web/packages/tidyverse/index.html): for non-spatial data handling

-   [**sfdep**](https://sfdep.josiahparry.com/): for spatial analysis

-   [**knitr**](https://cran.r-project.org/web/packages/knitr/):for prettifying presentation

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse, knitr)
```

## Environment settings

We will also set the default settings on for this document

-   `tmap_style` to **natural**: for displaying the maps with preferred style

-   set **seed** for reproducibility of results

```{r}
tmap_style("natural")
set.seed(1234)
```

## Loading the data

::: callout-important
Before running this part, please run all the code chunks in [Data Wrangling](/Take-home_Ex1/Take-home_Ex1A.qmd) as it generates the data needed for this document.
:::

Use `read_rds()` to load the `rds` data needed for geovisualization and analysis.

```{r}
sg_honeycomb <- read_rds("data/rds/sg_honeycomb_with_num_busstops.rds")
bus_peaks_hc <- read_rds("data/rds/bus_peaks_hc202310.rds") %>%
  filter(NUM_BUS_STOPS > 0)
```

-   `sg_honeycomb` - contains the honeycomb grid that covers Singapore boundaries
-   `bus_peaks_hc` - bus commuter traffic data for peak times for **October 2023**

::: callout-important
As mentioned in [Geovisualization and Analysis](/Take-home_Ex1/Take-home_Ex1B.html#conclusion), we are only focusing on **October 2023** data as we observed the same patterns in the August and September.
:::

# Methodology

Local Indicator of Spatial Association (LISA) detects spatial clusters present in our geospatial data and also helps identify spatial outliers.

We will use this analysis method to identify which areas have the highest bus commuters and which ones have the least.

To perform a LISA Analysis, we will do the following:

-   Deriving contiguity weights using Queen's method

-   Computing global Moran's I

-   Computing local Moran's I

-   Visualizing local Moran's I and p-value

-   Visualizing LISA Map

-   Interpreting the results

We will do this analysis for the all the peak hours.

::: callout-important
For all analysis in this study, we will set the significance level, or $\alpha$ value to **0.05**.
:::

# Deriving contiguity weights using Queens' method

Before getting Moran's I, we need to get the contiguity weights first. We will use the Queen's method.

::: callout-note
As our cells are hexagons, they touch the neighbors via edges, not single points. Hence, both Rook and Queen's method will return the same results.
:::

We will use `st_contiguity()` and `st_contiguity()` to get the contiguity weights. We will set `allow_zero = TRUE` as we have *island hexagons* in our data set.

```{r}
wm_q <-
  bus_peaks_hc %>%
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb,
                    style = "W",
                    allow_zero = TRUE)
  )
glimpse(wm_q)
```

# Calculating global Moran's I

To check for signs of clustering, we will first calculate the global Moran's I value.

-   When $I > 0$, observations are similar (sign of clustering)

-   When $I < 0$, observations are dissimilar (low indication of clustering)

-   If $p < \alpha$ (**0.05**)**,** the result is significant and **did not happen by chance**

We will do this for all peak times.

To perform the permutation test, we will use `global_moran_perm()`.

::: panel-tabset
## Weekday morning

```{r}
global_moran_perm(
  wm_q$WEEKDAY_AM_TRIPS,
  wm_q$nb,
  wm_q$wt,
  nsim = 99,
  zero.policy = TRUE
)
```

-   $0.23237 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

**Conclusion:** There are signs of geospatial clusters in the data.

## Weekday evening

```{r}
global_moran_perm(
  wm_q$WEEKDAY_PM_TRIPS,
  wm_q$nb,
  wm_q$wt,
  nsim = 99,
  zero.policy = TRUE
)
```

-   $0.06418 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

**Conclusion:** There are signs of geospatial clusters in the data.

## Weekend morning

```{r}
global_moran_perm(
  wm_q$WEEKEND_AM_TRIPS,
  wm_q$nb,
  wm_q$wt,
  nsim = 99,
  zero.policy = TRUE
)
```

-   $0.17209 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

**Conclusion:** There are signs of geospatial clusters in the data.

## 

## Weekend evening

```{r}
global_moran_perm(
  wm_q$WEEKEND_PM_TRIPS,
  wm_q$nb,
  wm_q$wt,
  nsim = 99,
  zero.policy = TRUE
)
```

-   $0.11845 > 0$ means there are signs of clustering

-   $p-value < 2.2\times10^{-16} < 0.05$ means the result is significant

**Conclusion:** There are signs of geospatial clusters in the data.
:::

# Calculating local Moran's I

Now that we have identified signs of clustering, we will identify where those clusters are.

We will do this for all peak times as clusters where detected for all

Before running the permutation test, we must set the seed first:

```{r}
set.seed(1234)
```

To perform the permutation test, we will use `global_moran_perm()`. We need to set `zero.policy = TRUE` as there are island hexagons in our data.

::: panel-tabset
## Weekday morning

```{r}
lisa_weekday_am <- wm_q %>% 
  mutate(local_moran = local_moran(
    WEEKDAY_AM_TRIPS, nb, wt, nsim = 99, zero.policy = TRUE),
         .before = 1) %>%
  unnest(local_moran)
kable(head(lisa_weekday_am))
```

## Weekday evening

```{r}
lisa_weekday_pm <- wm_q %>% 
  mutate(local_moran = local_moran(
    WEEKDAY_PM_TRIPS, nb, wt, nsim = 99, zero.policy = TRUE),
         .before = 1) %>%
  unnest(local_moran)
kable(head(lisa_weekday_pm))
```

## Weekend morning

```{r}
lisa_weekend_am <- wm_q %>% 
  mutate(local_moran = local_moran(
    WEEKEND_AM_TRIPS, nb, wt, nsim = 99, zero.policy = TRUE),
         .before = 1) %>%
  unnest(local_moran)
kable(head(lisa_weekend_am))
```

## Weekend evening

```{r}
lisa_weekend_pm <- wm_q %>% 
  mutate(local_moran = local_moran(
    WEEKEND_PM_TRIPS, nb, wt, nsim = 99, zero.policy = TRUE),
         .before = 1) %>%
  unnest(local_moran)
kable(head(lisa_weekend_pm))
```
:::

# Plotting local Moran's I and p-value

Next we will plot the local Moran's I values and p-values to gain insights.

::: panel-tabset
## Weekday morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
ii_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekday_am) +
    tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "local Moran's I of WEEKDAY_AM_TRIPS",
            main.title.size = 0.8)

p_ii_sim_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekday_am) +
  tm_fill("p_ii_sim",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
          labels = c("0.001", "0.01", "0.05", "Not sig"),
          palette = "YlOrRd"
          ) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(ii_map, p_ii_sim_map, ncol = 2)
```

Most of the map is covered by dark reds, corresponding to **insignificant values**. However, we can see the biggest cluster of **lighter hexagons on the West part**. Some lighter clusters were also seen in the map.

This peak time also has the **widest range** **of ii** values (**-5 to 25**), indicating that values are farthest from each other.

Plotting the LISA map later would reveal where the clusters are located based on `WEEKDAY_AM_TRIPS`.

## Weekday evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
ii_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekday_pm) +
    tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "local Moran's I of WEEKDAY_PM_TRIPS",
            main.title.size = 0.8)

p_ii_sim_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekday_pm) +
  tm_fill("p_ii_sim",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
          labels = c("0.001", "0.01", "0.05", "Not sig"),
          palette = "YlOrRd"
          ) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(ii_map, p_ii_sim_map, ncol = 2)
```

Similar to the **weekday evening** maps, most of the map is covered by dark reds, corresponding to **insignificant values**. However, there appears to be much more dark red hexagons compared to weekday morning.

The global Moran's I value is **0.06418** is closer to 0 than weekday morning's value of **0.23237**, indicating the **clustering is weaker** for weekday evening. This peak time also has the **narrowest range** **of ii** values (**-4 to 6**), indicating that values are closest to each other. This is consistent with our observation.

Plotting the LISA map later would reveal if there are indeed fewer clusters based on `WEEKDAY_PM_TRIPS` than `WEEKDAY_AM_TRIPS`.

## Weekend morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
ii_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekend_am) +
    tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "local Moran's I of WEEKEND_AM_TRIPS",
            main.title.size = 0.8)

p_ii_sim_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekend_am) +
  tm_fill("p_ii_sim",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
          labels = c("0.001", "0.01", "0.05", "Not sig"),
          palette = "YlOrRd"
          ) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(ii_map, p_ii_sim_map, ncol = 2)
```

The `p_ii_sim` map looks very similar to the **weekend evening** map. It appears that the weekend peak patterns are similar throughout the day. This seems to support our earlier hypothesis that the bus commuter traffic more distributed throughout the day on weekends as people have more control of their schedule as there is no work or school.

## Weekend evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
ii_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekend_pm) +
    tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "local Moran's I of WEEKEND_PM_TRIPS",
            main.title.size = 0.8)

p_ii_sim_map <-
  tm_shape(sg_honeycomb) +
  tm_fill(col="white") +
  tm_shape(lisa_weekend_pm) +
  tm_fill("p_ii_sim",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
          labels = c("0.001", "0.01", "0.05", "Not sig"),
          palette = "YlOrRd"
          ) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(ii_map, p_ii_sim_map, ncol = 2)
```

The `p_ii_sim` map looks very similar to the **weekend morning** map. It appears that the weekend peak patterns are similar throughout the day. This seems to support our earlier hypothesis that the bus commuter traffic more distributed throughout the day on weekends as people have more control of their schedule as there is no work or school.
:::

Observations from this part:

-   Indication of clusters at the West side

-   Highest indication of clustering for **weekday morning** peak times, consistent with its highest global Moran's I value

-   Similarity in clusters between weekend morning and evening peak times

# Plotting LISA maps

We can finally plot the LISA Maps to reveal where the clusters and outliers are.

::: panel-tabset
## Weekday morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
lisa_sig <- lisa_weekday_am  %>%
  filter(p_ii_sim < 0.05)

tm_shape(sg_honeycomb) +
  tm_fill(col="white")+
  tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4) +
  tm_layout(
    main.title = "LISA for Weekday 6 - 9AM",
    main.title.position = "center",
    main.title.size = 1
  )
```

Bigger **high-high** clusters around bus interchanges. There is high demand for buses in these areas in the morning, possibly being around residential areas.

## Weekday evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
lisa_sig <- lisa_weekday_pm  %>%
  filter(p_ii_sim < 0.05)

tm_shape(sg_honeycomb) +
  tm_fill(col="white")+
  tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4) +
  tm_layout(
    main.title = "LISA for Weekday 5 - 8PM",
    main.title.position = "center",
    main.title.size = 1
  )
```

Clusters are sparser compared to **weekday morning**, supporting our previous hypothesis that people going home from work or school are spread out throughout the night. This is because of after work activities (hanging out with friends, gym, shopping, etc.)

## Weekend morning

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
lisa_sig <- lisa_weekend_am  %>%
  filter(p_ii_sim < 0.05)

tm_shape(sg_honeycomb) +
  tm_fill(col="white")+
  tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4) +
  tm_layout(
    main.title = "LISA for Weekend 11AM - 2PM",
    main.title.position = "center",
    main.title.size = 1
  )
```

There are some **high-high** cells observed around the Central and South area, seemingly around the Clarke Quay, Mt. Faber and Sentosa areas.

Aside from this, there are singular **high-high** cells observed sparsely.

This could be because these are popular places for recreation that people go to unwind during weekends and holidays.

## Weekend evening

```{r}
#| code-fold: true
#| code-summary: "**Show the code**"
lisa_sig <- lisa_weekend_pm  %>%
  filter(p_ii_sim < 0.05)

tm_shape(sg_honeycomb) +
  tm_fill(col="white")+
  tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4) +
  tm_layout(
    main.title = "LISA for Weekend 4 - 7PM",
    main.title.position = "center",
    main.title.size = 1
  )
```

There are some **high-high** cells observed around the Central and South area, seemingly around the Clarke Quay and Bugis areas.

Aside from this, there are singular **high-high** cells observed sparsely.

This could be because these are popular places for late-afternoon and night activities during weekends, i.e. there are a lot of bars and restaurants around Clarke Quay and Bugis.
:::

## General observations

-   **Low-low** clusters in the West part (Jurong area), indicating **low demand for buses**.

    -   Are alternative modes of transport (MRT, taxi, car, etc) more popular in this area?

    -   Is this an area with sparse population?

-   **High-high** clusters seem to correspond to bus interchanges.

    -   *Eyeballed the map based on bus interchanges from* [Google Maps](https://www.google.com/maps/search/singapore+bus+interchange/@1.3385457,103.6866741,14.49z?entry=ttu)

    -   Multiple bus lines start at interchanges so people are expected to ride here, especially as they are close to residential areas

-   **Woodlands Checkpoint** is in a **high-high** cell

    -   People crossing the border into Singapore tap in here. Workers that reside in Johor Bahru utilize the bus stop here.

-   **Low-high** cells observed around **high-high** cells

    -   Indication under-utilized stops as people prefer to ride buses in nearby bus stops
