---
title: "In-class Ex 5: Spatial Economic Interaction Models"
author: "Kristine Joy Paas"
date: "16 December 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: false
  warning: false
---

# Setup Environment

-   [**spflow**](https://lukece.github.io/spflow/): R package for Spatial Economic Interaction Models

::: callout-important
We will the development version of `spflow`, which has major updates to syntax. (use at own risk)

It is not yet available on <https://cran.r-project.org/>, so we will download it from Github.
:::

```{r}
#| label: setup
devtools::install_github("LukeCe/spflow")
pacman::p_load(tmap, sf, spdep, sp, Matrix, knitr, tidyverse)
tmap_mode("plot")
tmap_style("natural")
```

# Loading data

```{r}
mpsz_nb <- read_rds("data/rds/mpsz_nb.rds")
mpsz_flow <- read_rds("data/rds/mpsz_flow.rds")
mpsz_var <- read_rds("data/rds/mpsz_var.rds")
```

For our model, we will use contiguity-based neighborhood structure.

# Preparing spflow objects

## Creating spflow_network

```{r}
mpsz_net <- spflow_network(
  id_net = "sg",
  node_neighborhood =
    nb2mat(mpsz_nb$by_contiguity),
  node_data = mpsz_var,
  node_key_column = "SZ_CODE")

mpsz_net
)
```

## Creating spflow networks pairs

```{r}
mpsz_net_pairs <- spflow_network_pair(
  id_orig_net = "sg",
  id_dest_net = "sg",
  pair_data = mpsz_flow,
  orig_key_column = "ORIGIN_SZ",
  dest_key_column = "DESTIN_SZ"
)

mpsz_net_pairs
```

## Creating spflow multi-network

::: callout-important
Output must be N by N, like the dimensions of `mpsz_flow`

```{r}
nrow(mpsz_var)^2 == nrow(mpsz_flow)
```
:::

```{r}
mpsz_multi_net <- spflow_network_multi(mpsz_net,
                                       mpsz_net_pairs)
mpsz_multi_net
```

## Correlation Analysis

Avoid having more than 2 explanatory variables that are linearly correlated. Otherwise, there will be bias in data. This is why it is important to generate a **correlogram** or **correlation matrix** to check which explanatory variables are correlated.

`P_` refers to impedance

```{r}
cor_formula <- log(1 + TRIPS) ~
  BUSSTOP_COUNT +
  AGE7_12 +
  AGE13_24 +
  AGE25_64 +
  SCHOOL_COUNT +
  BUSINESS_COUNT +
  RETAILS_COUNT +
  FINSERV_COUNT +
  P_(log(DISTANCE + 1))

cor_mat <- pair_cor(
  mpsz_multi_net,
  sp_flow_formula = cor_formula,
  add_lags_x = FALSE
)

colnames(cor_mat) <- paste0(
  substr(
    colnames(cor_mat), 1, 3), "...")

cor_image(cor_mat)
```

# Model Calibration
