---
title: "Hands-on Exercise 2C: Local Measures of Spatial Autocorrelation"
author: "Kristine Joy Paas"
date: "22 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

This hands-on exercise covers [Chapter 10: Local Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10)

I learned about the following:

-   Global Spatial Autocorrelation (GSA) statistics
-   Local Indicator of Spatial Association (LISA) statistics
-   Getis-Ord's Gi-statistics

## Preparing the data sets

Data sets used on this exercise were downloaded from [E-learn](https://elearn.smu.edu.sg/d2l/home/357628).

### Geospatial

-   Hunan county boundary layer (`shp` format)

### Aspatial

-   Hunan's local development indicators in 2012 (`csv` format)

Next, is putting them under the `Hands-on_Ex2` directory, with the following file structure:

``` bash
Hands-on_Ex2
└── data
    ├── aspatial
    │   └── Hunan_2012.csv
    └── geospatial
        ├── Hunan.dbf
        ├── Hunan.prj
        ├── Hunan.qpj
        ├── Hunan.shp
        └── Hunan.shx
```

## Installing R packages

I used the code below to install the R packages used in the exercise:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

# Getting the Data Into R Environment

::: callout-important
The steps here are similar to [Hands-on Exercise 2B's Getting the Data Into R Environment](/Hands-on_Ex2/Hands-on_Ex2B.html#getting-the-data-into-r-environment)

However, I copied all steps here so this page can run all the R code by itself.
:::

## Importing data sets

I used `st_read()` to import the geospatial `shp` data.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

::: callout-note
In the previous exercises, we transformed the data with **EPSG:3414**. However, that is not applicable for this data set as we are not working with Singapore 🇸🇬 data set.
:::

As with the previous exercises, I used `read_csv()` to import aspatial `csv` data.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Joining the data sets

In the exercise, we have to join the 2 data sets using this code:

```{r}
hunan <- left_join(hunan, hunan2012)%>%
  select(1:4, 7, 15)
```

::: callout-note
We did not specify any columns to join by but `left_join` detected common column, `County`, so it joined the 2 data sets by this column.

At the end of this, we are left with 7 columns, which includes **GDPPC** from the aspatial data, which contains data for **Gross Domestic Product per Capita**.
:::

## Visualizing Regional Development Indicator

Next, I plotted the GDPPC maps using **equal interval classification** and **equal quantile classification**.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

# Global Spatial Autocorrelation

::: callout-important
The steps here are similar to [Hands-on Exercise 2B's Global Spatial Autocorrelation](/Hands-on_Ex2/Hands-on_Ex2B.html#global-spatial-autocorrelation)

However, I copied all steps here so this page can run all the R code by itself.
:::

## Computing Contiguity Spatial Weights

First, I built the neighbor list using Queen contiguity-based neighbors. This means the regions must share a border (minimum a point) to be considered neighbors.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

## Row-standardized weights matrix

Next, I assigned weights to each neighboring county with value `1/(# of neighbors)`. This could be done by using `style="W"` to `nb2listw()`.

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

## Global Spatial Autocorrelation: Moran's I

### Moran's I test {#morans-i-test}

Next, I used Moran's I statistical testing using `moran.test()`.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

::: callout-tip
The Moran's I value is **0.30075**, which is greater than 0. This means that observations are **clustered**, and tend to be similar.

The p-value is also very close to 0, which indicates **high confidence on the correlation**.
:::

### Computing Monte Carlo Moran's I

Next, a Monte Carlo simulation was performed for the Moran's I statistic. 1000 simulations were performed by the code below:

```{r}
set.seed(1234)
bperm = moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

::: callout-tip
The Moran's I value is **0.30075**, same result as that of the [Moran's I test](#morans-i-test). Similarly, it means that observations are **clustered**, and tend to be similar.

The p-value is also very close to 0, which indicates **high confidence on the correlation**.
:::

### Visualizing Monte Carlo Moran's I

First, I examined the statistics of the Monte Carlo Moran's I. I checked the **mean**, **variance**, and the quantiles.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

Next, I also plotted the histogram.

::: callout-important
I plotted using **ggplot2** as an additional challenge from the exercise.
:::

```{r}
mc_results_df <- data.frame(moran_i = bperm$res)
ggplot(mc_results_df, aes(x = moran_i)) +
  geom_histogram(bins = 20, fill = "grey", color = "black") +
  geom_vline(xintercept = 0, color = "red") +
  labs(x = "Sumilated Moran's I",
       y = "Frequency")
```

::: callout-note
From this Monte Carlo simulations, results are skewed to the left, meaning most of the Moran's I simulations result in negative values. It means that in most simulation results, there is dispersion so there is **no spatial correlation**.

This is quite contradictory to the statistic from `moran.test`.

However, as this is a simulation set using seed **1234**, results could be different in other simulations because the sampling is different.
:::

## Global Spatial Autocorrelation: Geary's

Next I used Geary's method for spatial correlation.

### Geary's C test {#gearys-c-test}

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

::: callout-tip
Geary's C statistic is **0.6907**, which is **less than 1**. This means that observations are **clustered**, and tend to be similar. P-value is also very close to 0, suggesting high-confidence.\
\
It is consistent with the conclusions in [Moran's I test](#morans-i-test).
:::

### Computing Monte Carlo Geary's C

Similarly, I did permutation test via Monte Carlo simulations.

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

::: callout-tip
The Moran's I value is **0.6907**, same result as that of the [Geary's C test](#gearys-c-test). Similarly, it means that observations are **clustered**, and tend to be similar.

The p-value is also very close to 0, which indicates **high confidence on the correlation**.
:::

### Visualizing Monte Carlo Geary's C

First, I examined the statistics of the Monte Carlo Geary's C. I checked the **mean**, **variance**, and the quantiles.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

Finally, visualizing it.

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red")
```

::: callout-note
From this Monte Carlo simulations, results are quite balanced on 1, which makes it inconclusive as to the spatial clustering and dispersion.

This is quite contrary to the statistic resulting from `geary.test()`, which was more conclusive.

However, as this is a simulation set using seed **1234**, results could be different in other simulations because the sampling is different.
:::

# Spatial Correlograms

::: callout-important
The steps here are similar to [Hands-on Exercise 2B's Spatial Correlograms](/Hands-on_Ex2/Hands-on_Ex2B.html#spatial-correlograms)

However, I copied all steps here so this page can run all the R code by itself.
:::

Next, was to generate correlograms to look at patterns from a different perspective. This is done via `st.correlogram()`.

## Compute Moran's I correlogram {#compute-morans-i-correlogram}

First, I generated the correlogram for Morans's I.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

This did not provide me much information and I didn't know how to interpret it so I printed the full result.

```{r}
print(MI_corr)
```

::: callout-tip
From my understanding, since Moran's I values are greater than 0 and highest on lag 1, it means that the spatial correlation is most significant the closer the regions are.
:::

## Compute Geary's C correlogram

Next, I generated the correlogram for Geary's C.

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

Next was to print the results.

```{r}
print(GC_corr)
```

::: callout-tip
The Geary's C values are closest to 0 on the lag distance 1. Similar to [Compute Moran's I correlogram](#compute-morans-i-correlogram), the spatial correlation is strongest the closer the regions are.

The pattern is inverse of the Moran's I correlogram, which makes sense as Moran's I and Geary's C trends are inverse of each other.
:::

# Cluster and Outlier Analysis

The previous chapters are about Global Spatial Autocorrelation. In this part, I looked for local patterns that occur in subsets of the geospatial data.

## Computing Local Moran's I

First, I started with computing local Moran's I values.

The code chunks below were used to compute local Moran's I of *GDPPC2012* at the county level.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

This code chunk result in a matrix with columns:

-   Ii: the local Moran's I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

The local Moran's I values were inspected by:

```{r}
printCoefmat(data.frame(
  localMI[fips,], 
  row.names=hunan$County[fips]),
  check.names=FALSE)
```

## Mapping local Moran's I

Before proceeding with the mapping, I appended `localMI` dataframe onto the `hunan` dataframe.

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

Then, I plotted a cloropeth map of the local Moran's I values and the p-values using `tmap` functions. These maps were plotted side by side for easier analysis.

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

# Creating a LISA Cluster Map

The LISA Cluster Map shows the significant locations color coded by type of spatial autocorrelation.

## Plotting Moran scatterplot

In order to do this, I have to plot the Moran's I scatterplot first. This can be via `moran.plot()`.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

::: callout-note
This can be be interpreted such that the counties on the upper-right quadrant (e.g., **Shaosan**, **Ningxian**, **Liuyang**, **Wangchen**, **Changsa**) are within an affluent region, i.e., cluster of counties with high GDP per capita.

Some other counties of interest are **Zixing** and **Lengshuijiang**, which are more affluent than their neighbors. Lastly, **Pingjian** is less affluent compared to its neighbors.
:::

## Plotting Moran scatterplot with standardized variable

Next is to scale the plot by normalizing the axes, which should align the axes to `0`. `scale()` was used for this purpose.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% 
  as.vector 
```

After scaling, I replotted the scatterplot.

```{r}
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

## Preparing LISA map classes

To prepare LISA cluster map, I had to first create a numeric vector with the same number of elements as `localMI`, which is `88`.

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
```

Next was to compute the lag values and centering on the `mean`.

```{r}
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)
```

Similarly, i also centered the local Moran's I values around the mean.

```{r}
LM_I <- localMI[,1] - mean(localMI[,1])    
```

I also set the significant value to `0.05` as per standards.

```{r}
signif <- 0.05
```

Then, I defined the low-low (1), low-high (2), high-low (3) and high-high (4) categories. This corresponds to the quadrants in the scatterplot from [Plotting Moran scatterplot].

```{r}
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4
```

Lastly, was to add a category for non-significant Moran's I values.

```{r}
quadrant[localMI[,5]>signif] <- 0
```

## Plotting LISA map

After preparing the classes, I could finally plot the LISA map. As with the other maps so far, I used `tmap()` functions to created this map.

For easier analysis, I plotted the LISA map next to the GDPPC map.

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, 
             asp=1, ncol=2)
```

::: callout-note
I was expecting the 2 standalone orange counties from the GDPPC map (**Zixing** and **Lengshuijiang**) to be part of the high-low category. This is because they are relatively more affluent than their neighbors.

They were also on the high-low quadrant in the scatterplot. Hence, this result was surprising for me.

A possible explanation for this is that their GDPPC are just **a little bit higher than 60,000**, while their neighbors are in the high **50,000s**. Visually, they distinct but a closer look at the number might reveal that the values are not really far-off.
:::

I also plotted the local Moran's I values and p-values side by side again to find clues as to why.

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

::: callout-note
As I mentioned, **Zixing** and **Lengshuijiang** were part of the high-low category as I originally expected.

The p-value provides a sound explanation why. This is because the p-values for these counties are **0.100 or more**, which is more than the significance value that was set, which was **0.05**.

We can say that the p-value map can be use as a filter such that those counties with p-values greater than the significance value are considered **insignificant**, and only those are not included in this group will be categorized.
:::

# Hot Spot and Cold Spot Area Analysis

Beside detecting cluster and outliers, localised spatial statistics can be also used to detect hot spot and/or cold spot areas.

::: callout-tip
In the current exercise, these are what the terminologies mean:

**Hot spot** - counties with higher GDPPC compared to its surroundings

**Cold spot** - counties with lower GDPPC compares to its surroundings
:::

The analysis was done with Getis and Ord's G-statistics and consists of 3 steps:

-   Deriving spatial weight matrix

-   Computing Gi statistics

-   Mapping Gi statistics

## Deriving spatial weight matrix

::: callout-important
The code chunks used in this part are the same as the ones used in [Hands-on Exercise 2A: Spatial Weights and Applications](/Hands-on_Ex2/Hands-on_Ex2A.html). I didn't dive deep into these part as these was already learned.

For this exercise, **binary spatial weights** are used.
:::

In order to calculate the spatial weights, I needed to get determine the cut-off distance first. This was done by deriving the centroids and calculating the distances to the nearest neighbor for each county.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
coords <- cbind(longitude, latitude)
head(coords)
```

These are the same steps as in \[Hands-on Exercise 2A: Spatial Weights and Applications\](/Hands-on_Ex2/Hands-on_Ex2A.html#determining-cut-off-distance), where we determined the cut-off distance to be **62km**.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

Finally, the spatial weights matrix can be generated.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

As in the previous exercise, we could standardize the number of neighbors. This is because denser areas have more neighbors, while rural areas have less.

The code chunks below demonstrates how to standardize to 8 neighbors.

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

## Gi statistics with using fixed distance weights

Continuing from the steps above, I looked at two cases. In this part, I used the fixed distance weights.

First, I calculated the Gi statistics using the fixed distance weights.

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

Next is to add the Gi statistics to the `hunan` data frame.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

Finally, we could map the Gi values.

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

::: callout-note
There is a cluster of counties with high GDDPC in the Eastern part of China.

A striking observation is that the other orange cities outside of this cluster are **cold** in the local Gi map. This means that they are surrounded by counties with low GDPPC.

There should be caution when interpreting the map on the right as it is not intuitive because the values are actually based on the neighbors, and not the counties themselves.
:::

## Gi statistics with using adaptive distance weights

Next we calculate the Gi statistics using adaptive distance weights.

The steps are the same as in [Gi statistics with using fixed distance weights] but instead using the adaptive weights (`knn_lw`) instead of the fixed weights(`wm62_lw`).

First was to calculate the Gi statistics.

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
gi.adaptive
```

Then attaching the Gi statistics to the `hunan` data frame.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

And finally, mapping it.

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```

::: callout-note
With this plot, the clusters are a lot more clearer and we can clearly see that the East side is the **hot spot** in terms on GDDPC.

However, we can also notice that on the Southwest, the most affluent county is a **cold spot** in the Gi map because it is surrounded by less affluent counties.
:::

Another observation is that the Gi map using adaptive distance weights are *less scattered* and bigger compared to the Gi map using fixed distance weights.

# Reflections

From this exercise, I realized how maps (and graphs) can be misleading. Difference in colors in the maps made me intuitively think that 2 counties have much different GDPPC values compared to their neighbors, when in reality, the difference is actually insignificant.

For example, if 0-99 is colored as light orange while 100-199 is orange, a region in the map with value of 99 can look much different than its neighbor with value 100. In contrast, another neighbor with value 50 would look similar despite its value being much farther to 99 than 100 (which is actually insignificant).

This is dangerous in a way that data visualization can be weaponized. These boundaries between values can be manipulated to fit certain narratives. I just realized that even a relatively data literate person can me can be fooled by this and this makes me scared, given the misinformation in the digital age.

From now on, I will do more diligence in verifying the data presented to me.
