---
title: "Hands-on Exercisa 2B: Global Measures of Spatial Autocorrelation"
author: "Kristine Joy Paas"
date: "20 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

This hands-on exercise covers [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09)

I learned about the following:

-   Spatial correlation
-   Global Spatial Autocorrelation (GSA) statistics
-   Local Indicator of Spatial Association (LISA) statistics
-   Getis-Ord\'s Gi-statistics

# Getting Started

## Preparing the data sets

Data sets used on this exercise were downloaded from [E-learn](https://elearn.smu.edu.sg/d2l/home/357628).

### Geospatial

-   Hunan county boundary layer (`shp` format)

### Aspatial

-   Hunan's local development indicators in 2012 (`csv` format)

Next, is putting them under the `Hands-on_Ex2` directory, with the following file structure:

``` bash
Hands-on_Ex2
└── data
    ├── aspatial
    │   └── Hunan_2012.csv
    └── geospatial
        ├── Hunan.dbf
        ├── Hunan.prj
        ├── Hunan.qpj
        ├── Hunan.shp
        └── Hunan.shx
```

## Installing R packages

I used the code below to install the R packages used in the exercise:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

# Getting the Data Into R Environment\*\*

## Importing data sets

I used `st_read()` to import the geospatial `shp` data.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

::: callout-note
In the previous exercises, we transformed the data with **EPSG:3414**. However, that is not applicable for this data set as we are not working with Singapore 🇸🇬 data set.
:::

As with the previous exercises, I used `read_csv()` to import aspatial `csv` data.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Joining the data sets

In the exercise, we have to join the 2 data sets using this code:

```{r}
hunan <- left_join(hunan, hunan2012)%>%
  select(1:4, 7, 15)
```

::: callout-note
We did not specify any columns to join by but `left_join` detected common column, `County`, so it joined the 2 data sets by this column.

At the end of this, we are left with 7 columns, which includes **GDPPC** from the aspatial data, which contains data for **Gross Domestic Product per Capita**.
:::

# Reflections

*Reflections go here*
