---
title: "Hands-on Exercisa 2B: Global Measures of Spatial Autocorrelation"
author: "Kristine Joy Paas"
date: "20 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

This hands-on exercise covers [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09)

I learned about the following:

-   Spatial correlation using Moran's I and Geary's C
-   Correlograms

## Preparing the data sets

Data sets used on this exercise were downloaded from [E-learn](https://elearn.smu.edu.sg/d2l/home/357628).

### Geospatial

-   Hunan county boundary layer (`shp` format)

### Aspatial

-   Hunan's local development indicators in 2012 (`csv` format)

Next, is putting them under the `Hands-on_Ex2` directory, with the following file structure:

``` bash
Hands-on_Ex2
└── data
    ├── aspatial
    │   └── Hunan_2012.csv
    └── geospatial
        ├── Hunan.dbf
        ├── Hunan.prj
        ├── Hunan.qpj
        ├── Hunan.shp
        └── Hunan.shx
```

## Installing R packages

I used the code below to install the R packages used in the exercise:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

# Getting the Data Into R Environment

## Importing data sets

I used `st_read()` to import the geospatial `shp` data.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

::: callout-note
In the previous exercises, we transformed the data with **EPSG:3414**. However, that is not applicable for this data set as we are not working with Singapore 🇸🇬 data set.
:::

As with the previous exercises, I used `read_csv()` to import aspatial `csv` data.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Joining the data sets

In the exercise, we have to join the 2 data sets using this code:

```{r}
hunan <- left_join(hunan, hunan2012)%>%
  select(1:4, 7, 15)
```

::: callout-note
We did not specify any columns to join by but `left_join` detected common column, `County`, so it joined the 2 data sets by this column.

At the end of this, we are left with 7 columns, which includes **GDPPC** from the aspatial data, which contains data for **Gross Domestic Product per Capita**.
:::

## Visualizing Regional Development Indicator

Next, I plotted the GDPPC maps using **equal interval classification** and **equal quantile classification**.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

# Global Spatial Autocorrelation

## Computing Contiguity Spatial Weights

First, I built the neighbor list using Queen contiguity-based neighbors. This means the regions must share a border (minimum a point) to be considered neighbors.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

## Row-standardized weights matrix

Next, I assigned weights to each neighboring county with value `1/(# of neighbors)`. This could be done by using `style="W"` to `nb2listw()`.

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

## Global Spatial Autocorrelation: Moran's I

### Moran's I test {#morans-i-test}

Next, I used Moran's I statistical testing using `moran.test()`.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

::: callout-tip
The Moran's I value is **0.30075**, which is greater than 0. This means that observations are **clustered**, and tend to be similar.

The p-value is also very close to 0, which indicates **high confidence on the correlation**.
:::

### Computing Monte Carlo Moran's I

Next, a Monte Carlo simulation was performed for the Moran's I statistic. 1000 simulations were performed by the code below:

```{r}
set.seed(1234)
bperm = moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

::: callout-tip
The Moran's I value is **0.30075**, same result as that of the [Moran's I test](#morans-i-test). Similarly, it means that observations are **clustered**, and tend to be similar.

The p-value is also very close to 0, which indicates **high confidence on the correlation**.
:::

### Visualizing Monte Carlo Moran's I

First, I examined the statistics of the Monte Carlo Moran's I. I checked the **mean**, **variance**, and the quantiles.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

Next, I also plotted the histogram.

::: callout-important
I plotted using **ggplot2** as an additional challenge from the exercise.
:::

```{r}
mc_results_df <- data.frame(moran_i = bperm$res)
ggplot(mc_results_df, aes(x = moran_i)) +
  geom_histogram(bins = 20, fill = "grey", color = "black") +
  geom_vline(xintercept = 0, color = "red") +
  labs(x = "Sumilated Moran's I",
       y = "Frequency")
```

::: callout-note
From this Monte Carlo simulations, results are skewed to the left, meaning most of the Moran's I simulations result in negative values. It means that in most simulation results, there is dispersion so there is **no spatial correlation**.

This is quite contradictory to the statistic from `moran.test`.

However, as this is a simulation set using seed **1234**, results could be different in other simulations because the sampling is different.
:::

## Global Spatial Autocorrelation: Geary's

Next I used Geary's method for spatial correlation.

### Geary's C test {#gearys-c-test}

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

::: callout-tip
Geary's C statistic is **0.6907**, which is **less than 1**. This means that observations are **clustered**, and tend to be similar. P-value is also very close to 0, suggesting high-confidence.\
\
It is consistent with the conclusions in [Moran's I test](#morans-i-test).
:::

### Computing Monte Carlo Geary's C

Similarly, I did permutation test via Monte Carlo simulations.

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

::: callout-tip
The Moran's I value is **0.6907**, same result as that of the [Geary's C test](#gearys-c-test). Similarly, it means that observations are **clustered**, and tend to be similar.

The p-value is also very close to 0, which indicates **high confidence on the correlation**.
:::

### Visualizing Monte Carlo Geary's C

First, I examined the statistics of the Monte Carlo Geary's C. I checked the **mean**, **variance**, and the quantiles.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

Finally, visualizing it.

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red")
```

::: callout-note
From this Monte Carlo simulations, results are quite balanced on 1, which makes it inconclusive as to the spatial clustering and dispersion.

This is quite contrary to the statistic resulting from `geary.test()`, which was more conclusive.

However, as this is a simulation set using seed **1234**, results could be different in other simulations because the sampling is different.
:::

# Spatial Correlograms

Next, was to generate correlograms to look at patterns from a different perspective. This is done via `st.correlogram()`.

## Compute Moran's I correlogram {#compute-morans-i-correlogram}

First, I generated the correlogram for Morans's I.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

This did not provide me much information and I didn't know how to interpret it so I printed the full result.

```{r}
print(MI_corr)
```

::: callout-tip
From my understanding, since Moran's I values are greater than 0 and highest on lag 1, it means that the spatial correlation is most significant the closer the regions are.
:::

## Compute Geary's C correlogram

Next, I generated the correlogram for Geary's C.

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

Next was to print the results.

```{r}
print(GC_corr)
```

::: callout-tip
The Geary's C values are closest to 0 on the lag distance 1. Similar to [Compute Moran's I correlogram](#compute-morans-i-correlogram), the spatial correlation is strongest the closer the regions are.

The pattern is inverse of the Moran's I correlogram, which makes sense as Moran's I and Geary's C trends are inverse of each other.
:::

# Reflections

-   It took me a while to understand the implications of the Monte Carlo simulations with Moran's I and Geary's C. I spent a much longer time than expected on this shorter chapter.

-   Moran's I and Geary's C are inverse of each other and the patterns shown in this exercise look consistent. I wonder if it is worth knowing both or knowing just one is enough since they are for the same purpose.
