---
title: "In-class Exercise 1: Bus Commuter Flow by Origin"
author: "Kristine Joy Paas"
date: "18 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
---

# Overview

In this exercise, we will learn how to create choropleth map from bus data and commuter data.

# Getting Started

The code chunk below loads the following packages:

-   tmap: for thematic mapping

-   sf: for geospatial data handling

-   tidyverse: for non-spatial data handling

```{r}
pacman::p_load(tmap, sf, tidyverse)
```

# Preparing the Flow Data

## Importing the origin and destination data

Firstly, we will import the *Passenger Volume By Origin Destination Bust Stops* data set from LTA Data Mall by using `read_csv` .

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202308.csv")
```

To make it easier to process the PT codes, we will convert the origin and destination PT codes to factor data type.

```{r}
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
```

Next, I want to investigate the commuter flows between 7-10am on weekdays.

::: callout-tip
We need to use the interval `7 <= time <= 9` as the data with `time = 9` contains data from 9am to just before 10am.
:::

```{r}
origin7_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 7 &
           TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

Next, we save the data in `rds` format for future use.

::: callout-important
In order for the code below to work, we need to create the `rds/` directory under `data/`.
:::

```{r}
write_rds(origin7_9, "data/rds/origin7_9.rds")
```

It can be imported back later on with `read_rds()`.

```{r}
origin7_9 <- read_rds("data/rds/origin7_9.rds")
```

## Importing geospatial data

Next we need to import the bus stop locations so that we can correlate them from the PT codes from the origin and destination data from before.

```{r}
busstop <- st_read(dsn = "data/geospatial",
                  layer = "BusStop") %>%
  st_transform(crs=3414)
```

Since these are just points, we also need some polygon data to figure out where in the Singapore map the bus locations correspond to:

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

# Data Wrangling

After importing all the needed data into RStudio, we can proceed with data wrangling.

## Combining busstop and mpsz

We first combine that 2 data frames by figuring out which polygon in `mpsz` each points in `busstop` are contained in.

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>%
  select(BUS_STOP_N, SUBZONE_C) %>%
  st_drop_geometry()
glimpse(busstop_mpsz)
```

::: callout-note
-   `st_intersection()` is used to perform point and polygon overly and the output will be in point sf object.

-   `select()` of dplyr package is then use to retain only BUS_STOP_N and SUBZONE_C in the busstop_mpsz sf data frame.

-   `st_drop_geometry()` removes the geometry from the data frame, making the data aspatial.
:::

::: callout-tip
`busstop_mpsz` has **5 less** items than `busstop`. This is because these bus stops are outside of Singapore 🇸🇬 border.

For example, some bus routes start/end in Johor Bahru in Malaysia 🇲🇾.
:::

Finally, we now now which subzone each bus stop is located in.

Before proceeding, we should save the `busstop_mpsz` so we won't need to recalculate it later on.

```{r}
write_rds(busstop_mpsz, "data/rds/busstop_mpsz.csv")
```

Next is to join the `busstop_mpsz` data with `origin7_9`.

```{r}
origin_data <- left_join(origin7_9 , busstop_mpsz,
            by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE,
         ORIGIN_SZ = SUBZONE_C)
```

Next is to check for duplicate records.

```{r}
duplicate <- origin_data %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
glimpse(duplicate)
```

In this case, there are some duplicates so we need to clean it up further. This can be done using `unique()`.

```{r}
origin_data <- unique(origin_data)
```

Let's check if the duplicates have been removed.

```{r}
duplicate <- origin_data %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
glimpse(duplicate)
```

The duplicates have been removed so we can proceed with merging `origin_data` with `mpsz` to figure out the subzone names on the bus stop locations.

```{r}
mpsz_origtrip <- left_join(mpsz, 
                           origin_data,
                           by = c("SUBZONE_C" = "ORIGIN_SZ"))
```

Before proceeding, let's save this data as `rds`.

```{r}
write_rds(origin7_9, "data/rds/origin7_9.rds")
```

# Choropleth Mapping

Finally, we can generate a map for the bus trip origins.

```{r}
tm_shape(mpsz_origtrip)+
  tm_fill("TRIPS", 
          style = "quantile", 
          palette = "YlOrRd",
          title = "Passenger trips") +
  tm_layout(main.title = "Passenger trips started at sub-zones (Weekdays 7-10am)",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha = 0.1) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

We can also map it by region easier comparison.

```{r}
tm_shape(mpsz_origtrip) +
  tm_fill("TRIPS",
          style = "quantile",
          palette = "YlOrRd",
          thres.poly = 0,
          title = "Passenger trips") + 
  tm_facets(by="REGION_N",
            ncol = 2,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(main.title = "Passenger trips started at sub-zones (Weekdays 7-10am)",
            main.title.position = "center",
            main.title.size = 1.2,
            title.position = c("center", "center"), 
            title.size = 20,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.position = c("center", "bottom")) +
  tm_borders(alpha = 0.5)
```

::: callout-note
From these maps, we can infer that **Central Region** has the most bus trips started between 7-10am.

It makes sense as many business activities happen within this region so it is expected that many transport activities also happen. These activities can be attributed to workers taking a bus from their homes, or taking a bus for the final leg of their commute to the office, i.e., taking a train first then transferring to a bus.
:::

# Reflections

-   This exercise gives a good structure on how a basic geospatial data analysis is done. I want to explore the mapping more.
